An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-02-26 EconGrowth.Rnw
Doing some serious code cleanup. At this point, I have split the Cobb-Douglas functions apart, and everything is still working fine.

<<cobb-douglas functions, eval=TRUE>>=
# Returns an nls Cobb-Douglas model for the country specified
# Give an energyType ("Q", "X", or "U") if you want to include an energy term
# by including the gamma parameter, a fit with fixed gamma will be provided.
cobbDouglasModel <- function(countryAbbrev, energyType, gamma){
  dataTable <- loadData(countryAbbrev) #Load the data that we need.
  #Run the non-linear least squares fit to the data based on whether we have an energy type or not.
    # No energy term desired in the Cobb-Douglas equation.
    # Establish guess values for alpha and lambda.
    lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
    alphaGuess <- 0.7 #0.7 gives good results for all countries.  
    start <- list(lambda=lambdaGuess, alpha=alphaGuess)
    # Runs a non-linear least squares fit to the data. We've replaced beta with 1-alpha for simplicity.
    model <- iGDP ~ exp(lambda*iYear) * iCapStk^alpha * iLabor^(1 - alpha)
    modelCD <- nls(formula=model, data=dataTable, start = start,
                   #Include the next 3 lines to fit with constraints.
                   #algorithm = "port",
                   #lower = list(lambda=-Inf, alpha=0),
                   #upper = list(lambda=Inf, alpha=1)
    return(modelCD)
    # We need to do the Cobb-Douglas fit with the desired energy data.
    # To achieve the correct fit, we'll change the name of the desired column
    # to "iEToFit" and use "iEToFit" in the nls function. 
    dataTable <- replaceColName(dataTable, energyType, "iEToFit")
    # Reparameterize to ensure that we meet the constraints:
    # * alpha + beta + gamma = 1.0.
    # * alpha, beta, and gamma are all between 0.0 and 1.0.
    # To do this, we reparameterize as
    # * 0 < a < 1
    # * 0 < b < 1
    # * alpha = min(a, b)
    # * beta = b - a
    # * gamma = 1 - max(a, b)
    # Establish guess values for lambda, alpha, and beta.
    lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
    alphaGuess <- 0.9
    betaGuess <- 1.0 - alphaGuess
    if (countryAbbrev == "ZA"){
      # South Africa doesn't converge using the default guess values above. Try different values.
      gammaGuess <- 0.65 # Found via grid search over gamma
      alphaGuess <- 0.597 # The alpha value when energy is not included
      betaGuess <- 1.0 - alphaGuess - gammaGuess
#       alphaGuess <- 0.4
#       gammaGuess <- 0.1
#       betaGuess <- 1.0 - alphaGuess - gammaGuess
#       alphaGuess <- 0.39
#       betaGuess <- 0.41
    aGuess <- alphaGuess
    bGuess <- alphaGuess + betaGuess
    start <- list(lambda=lambdaGuess, a=aGuess, b=bGuess)
    # Now actually do the fit, using the column name "iEToFit".
    if (missing(gamma)){
      # gamma is a free parameter. We need to estimate lambda, a, and b.
      model <- iGDP ~ exp(lambda*iYear) * iCapStk^min(a,b) * iLabor^abs(b-a) * iEToFit^(1.0 - max(a,b))
      modelCDe <- nls(formula=model, data = dataTable, start = start,
                      control = nls.control(maxiter = 200, 
                                            tol = 1e-05, 
                                            minFactor = 1/1024, 
                                            printEval=FALSE, #Tells whether to print details of curve fit process.
                                            warnOnly=TRUE),
                      #Include the next 3 lines to fit with constraints.
                      #algorithm = "port",
                      #lower = list(lambda=-Inf, a=0, b=0),
                      #upper = list(lambda= Inf, a=1, b=1)
      # We don't want to estimate the value of gamma
      # Calculate a new column iEGamma = iEToFit^gamma
      iEGamma <- with(dataTable, iEToFit^gamma)
      dataTable <- cbind(dataTable, iEGamma)
      # We already know the value of gamma. We don't want to estimate it.
      # Establish guess values for lambda and alpha.
      lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
      alphaGuess <- 0.7*(1-gamma) #ensure a consistent starting point.
      start <- list(lambda=lambdaGuess, alpha=alphaGuess)
      model <- iGDP ~ exp(lambda*iYear) * iCapStk^alpha * iLabor^((1-gamma) - alpha) * iEGamma
      modelCDe <- nls(formula=model, data = dataTable, start = start,
                      control = nls.control(maxiter = 200, 
                                            tol = 1e-05, 
                                            minFactor = 1/1024, 
                                            printEval=FALSE, #Tells whether to print details of curve fit process.
                                            warnOnly=TRUE),
                      #Include the next 3 lines to fit with constraints.
                      algorithm = "port",
                      lower = list(lambda=-Inf, alpha=0),
                      upper = list(lambda=Inf, alpha=1.0-gamma)
# print(summary(modelCDe))
    return(modelCDe)

>>>>

<<cobb-douglas functions, eval=TRUE>>=
####################
# Returns an nls Cobb-Douglas model (without energy) for the country specified. 
# No energy is included in the function to be fitted.
##
cdModel <- function(countryAbbrev){
  dataTable <- loadData(countryAbbrev) #Load the data that we need.
  # Run the non-linear least squares fit to the data. No energy term desired in the Cobb-Douglas equation.
  # Establish guess values for alpha and lambda.
  lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
  alphaGuess <- 0.7 #0.7 gives good results for all countries.  
  start <- list(lambda=lambdaGuess, alpha=alphaGuess)
  # Runs a non-linear least squares fit to the data. We've replaced beta with 1-alpha for simplicity.
  model <- iGDP ~ exp(lambda*iYear) * iCapStk^alpha * iLabor^(1 - alpha)
  modelCD <- nls(formula=model, data=dataTable, start = start,
                 #Include the next 3 lines to fit with constraints.
                 #algorithm = "port",
                 #lower = list(lambda=-Inf, alpha=0),
                 #upper = list(lambda=Inf, alpha=1)
                 )
  return(modelCD)
}

####################
# Returns an nls Cobb-Douglas model for the country specified
# energyType is one of "Q", "X", or "U".
##
cdeModel <- function(countryAbbrev, energyType){
  dataTable <- loadData(countryAbbrev) #Load the data that we need.
  # We need to do the Cobb-Douglas fit with the desired energy data.
  # To achieve the correct fit, we'll change the name of the desired column
  # to "iEToFit" and use "iEToFit" in the nls function. 
  dataTable <- replaceColName(dataTable, energyType, "iEToFit")
  # Establish guess values for lambda, alpha, and beta.
  lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
  alphaGuess <- 0.9
  betaGuess <- 1.0 - alphaGuess
  aGuess <- alphaGuess
  bGuess <- alphaGuess + betaGuess
  start <- list(lambda=lambdaGuess, a=aGuess, b=bGuess)
  # Now actually do the fit, using the column name "iEToFit".
  # gamma is a free parameter.
  # Reparameterize to ensure that we meet the constraints:
  # * alpha + beta + gamma = 1.0.
  # * alpha, beta, and gamma are all between 0.0 and 1.0.
  # To do this, we reparameterize as
  # * 0 < a < 1
  # * 0 < b < 1
  # * alpha = min(a, b)
  # * beta = b - a
  # * gamma = 1 - max(a, b)
  model <- iGDP ~ exp(lambda*iYear) * iCapStk^min(a,b) * iLabor^abs(b-a) * iEToFit^(1.0 - max(a,b))
  modelCDe <- nls(formula=model, data = dataTable, start = start,
                  control = nls.control(maxiter = 200, 
                                        tol = 1e-05, 
                                        minFactor = 1/1024, 
                                        printEval=FALSE, #Tells whether to print details of curve fit process.
                                        warnOnly=TRUE),
                  #Include the next 3 lines to fit with constraints.
                  #algorithm = "port",
                  #lower = list(lambda=-Inf, a=0, b=0),
                  #upper = list(lambda= Inf, a=1, b=1)
                  )
  return(modelCDe)
}

##############################
# Returns a Cobb-Douglas model where gamma (the exponent on the energy term) has been fixed
##
cdeFixedGammaModel <- function(countryAbbrev, energyType, gamma){
  dataTable <- loadData(countryAbbrev) #Load the data that we need.
  # We need to do the Cobb-Douglas fit with the desired energy data.
  # To achieve the correct fit, we'll change the name of the desired column
  # to "iEToFit" and use "iEToFit" in the nls function. 
  dataTable <- replaceColName(dataTable, energyType, "iEToFit")
  # We're fixing the value of gamma.
  # Calculate a new column iEGamma = iEToFit^gamma
  iEGamma <- with(dataTable, iEToFit^gamma)
  dataTable <- cbind(dataTable, iEGamma)
  # We already know the value of gamma. We don't want to estimate it.
  # Establish guess values for lambda and alpha.
  lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
  alphaGuess <- 0.7*(1-gamma) #ensure a consistent starting point.
  start <- list(lambda=lambdaGuess, alpha=alphaGuess)
  model <- iGDP ~ exp(lambda*iYear) * iCapStk^alpha * iLabor^((1-gamma) - alpha) * iEGamma
  modelCDe <- nls(formula=model, data = dataTable, start = start,
                  control = nls.control(maxiter = 200, 
                                        tol = 1e-05, 
                                        minFactor = 1/1024, 
                                        printEval=FALSE, #Tells whether to print details of curve fit process.
                                        warnOnly=TRUE),
                  #Include the next 3 lines to fit with constraints.
                  algorithm = "port",
                  lower = list(lambda=-Inf, alpha=0),
                  upper = list(lambda=Inf, alpha=1.0-gamma)
                  )
  return(modelCDe)
}

####################
# Returns an nls Cobb-Douglas model for the country specified
# Give an energyType ("Q", "X", or "U") if you want to include an energy term
# If you supply a value for the gamma argument, a fit with fixed gamma will be provided.
# This function dispatches to cdModel, cdeModel, or cdFixedGammaModel as appropriate.
##
cobbDouglasModel <- function(countryAbbrev, energyType, gamma){
  if (missing(energyType)){
    # Fit the Cobb-Douglas model without energy.
    return(cdModel(countryAbbrev))
  } else {
    if (missing(gamma)){
      # Fit the Cobb-Douglas model with gamma as a free parameter
      return(cdeModel(countryAbbrev, energyType))
    }
    # Fit the Cobb-Douglas model with fixed value of gamma
    return(cdeFixedGammaModel(countryAbbrev, energyType, gamma))
  }
}
